Measuring causal influence with\\ back-to-back regression

### 2019-05-23 paper/introduction.tex

First, causal and non-causal factors may be numerous and collinear. In physics, for example, a student may have difficulty identifying whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factor increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity [CITE]. Indeed, word frequency strongly covaries with word length (short words are more frequent than long words) as well as with word category (determinants are more frequent than adverbs), the individual letters from which it is composed etc (e.g. words that contain "z" are rare) and so forth. As it can be particularly difficult to generate a set of words that control for all of these factors simultaneously, it is common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, etc), and investigate the resulting parameters of the best model [CITE].

The second challenge to measuring causal influence is that observations are becoming increasingly large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations. For example, brain activity is generally recorded with hundreds of sensors simultaneously. It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus [CITE].
%
Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors [CITE].

>>>>

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factor increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity as this factor covaries with numerous other factors such as word length (short words are more frequent than long words) word category (determinants are more frequent than adverbs), individual letters etc (e.g. words that contain "z" are rare) \cite{kutas2011thirty,pegado2014timing}. Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, etc, e.g. \cite{huth2016natural}), and investigate the resulting parameters of the best model.

The second challenge to measuring causal influence is that observations can be large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations. For example, brain activity is generally recorded with hundreds of sensors simultaneously. It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \cite{cichy2014resolving, king2016brain, kriegeskorte2008representational}.

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart_neuroimage_2015, king2018encoding}.

### 2019-05-23 paper/meg_exps.tex

% \paragraph{Forward regression} also known as 'encoding' in neuroimaging, here consists of
% fitting a ridge-regularized ordinary least squares (OLS) regression \footnote{We use scikit-learn \citep{sklearn}.} from $X$ to $Y$:
% \begin{equation}
%   H = (X^{T}X+\lambda I)^{-1} X^{T}Y
% \end{equation}
% where $\lambda$ is the
% regularization parameter, cross-validated between $10^{-5}$ and $10^5$.
%
% We then estimate the contribution
% $diag(\hat E) \in\mathbb{D}^{d} $ of each factor. However, $H\in\mathbb{R}^{d,
% q}$ is a matrix of $d$ word features by $q$ MEG channels. We approximate
% the relative contribution of each feature with the sum-of-squares of the corresponding coefficients in $H$; for the $j$-th feature:
% \begin{equation}
%   \hat e_j = \sum_{i=0}^{q} {h_{i,j}}^2
% \end{equation}
% \paragraph{Backward regression} or 'decoding', is similar but predicts $X$ from $Y$:
% \begin{equation} G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X \end{equation}
% where $\lambda$ is cross-validated as for encoding. We estimate the
% contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor in the following
% way. We predict $X$ on a held-out split \textit{test}:
% % Using 10-fold cross-validation, we thus assessed the ability of this ridge
% % regression to predict independent and identically distributed data:
% \begin{equation} \hat X_{test} = Y_{test}\cdot G_{train} \end{equation}
% and assess, for each factor of X independently, the Pearson correlation
% coefficient $r$ between $X_{test}$ and $\hat X_{test}$. We repeat the procedure
% for 20 random splits and average the result.

%<<<<<<< HEAD
% \paragraph{Back-to-back regression} We adapted back-to-back regression with two
% changes. First, to avoid getting a slightly positively-biased estimate of the
% factors contributing to the MEG signal, we swap the Tikhonov regularization of
% the first regression $G$ with a truncated regularization. Truncation was fitted
%  by maximum likelihood estimate, as implemented in scikit-learn principal
%  component analysis. Second, to minimize the variance increase induced by
%  bagging, we computed the inverse regularized covariance matrices outside
%  bagging:
% \begin{equation}
%   G = (Y^T Y+\lambda_Y I)^{-1} Y_{train}^T X_{train}
% \end{equation}
% \begin{equation}
% \hat X_{test} = Y_{test} \cdot G^T
% \end{equation} where $\left(train,test\right)$ are partitions of the data.

@@

The three types of analyses were implemented for each subject and each time
sample independently.  To estimate the reliability of these estimates in the
population, we tested, using one-sample $t$-test whether the distribution of
the estimated contribution of each factor different from baseline ($t=0$ms),
as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ ranging from
150ms to 1s after word onset. We use the resulting p-values as proxy to quantify
the performance of each model. These results are depicted in Figure
\ref{fig:meg_pval}. For concision purposes, we only report this metric averaged
between 150ms and 1s after word onset.
% =======
% \paragraph{Back-to-back regression} We adapted back-to-back regression with two
% changes. First, to avoid getting a slightly positively-biased estimate of the
% factors contributing to the MEG signal, we swap the Tikhonov regularization of
% the first regression $G$ with a truncated regularization. Truncation was fitted
%  by maximum likelihood estimate, as implemented in scikit-learn principal
%  component analysis. Second, to minimize the variance increase induced by
%  bagging, we computed the inverse regularized covariance matrices outside
%  bagging:
% \begin{equation}
%   G = (Y^T Y+\lambda_Y I)^{-1} Y_{train}^T X_{train}
% \end{equation}
% \begin{equation}
% \hat X_{test} = Y_{test} \cdot G^T
% \end{equation} where $\left(train,test\right)$ are partitions of the data.
% %
% % \subsubsection{Second-level statistics}
% %
% % The three types of analyses were implemented for each subject and each time
% % sample independently.  To estimate the reliability of these estimates in the
% % population, we tested, using one-sample $t$-test whether the distribution of
% % the estimated contribution of each factor different from baseline ($t=0$ms),
% % as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ 150ms and
% % 400ms respectively
% >>>>>>> 1775190a9737244f84b8ace529198a443029ad8d

@@

In Figure \ref{fig:meg_twocurves} we show, for each model, the effects obtained
in a representative subject, as well as the average effects across subjects.
The results show that all models significantly detect that both word length and
word frequency modulate brain responses after the stimulus onset. However
forward regression detected word length ($t-value=10.6$, $p=10^{-16}$) and
word frequency ($t-value=11.1$, $p=10^-17$) less reliably than B2B (length:
$t-value=18.7$, $p=10^{-31}, frequency:$t-value=15.9$, $p=10^{-26}$). B2B
was less sensitive than the backward regression (length: $t-value=33.6$,
$p=10^{-50}$, $t-value=31.7$, frequency:$p=10^{-48}$). However, as detailed in
the introduction, backward models confound collinear factors. Consequently,
the high sensitivity of our backward regression is likely to come from a loss of specificity. This possibility is corroborated by the fact that the dynamics
associated with word length and word frequency are remarkably similar in the
backward model.

>>>>

@@

The three types of analyses were implemented for each subject and each time
sample independently.  To estimate the reliability of these estimates in the
population, we tested, using one-sample $t$-test whether the distribution of
the estimated contribution of each factor different from baseline ($t=0$ms),
as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ ranging from
150ms to 1s after word onset. We use the resulting p-values as proxy to quantify
the performance of each model. These results are depicted in Figure
\ref{fig:meg_pval}. For concision purposes, we only report this metric averaged
between 150ms and 1s after word onset.

@@

In Figure \ref{fig:meg_twocurves} we show, for each model, the effects obtained
in a representative subject, as well as the average effects across subjects.
The results show that all models significantly detect that both word length and
word frequency modulate brain responses after the stimulus onset. However
forward regression detected word length ($t-value=10.6$, $p=10^{-16}$) and
word frequency ($t-value=11.1$, $p=10^-17$) less reliably than B2B (length:
$t-value=18.7$, $p=10^{-31}$, frequency:$t-value=15.9$, $p=10^{-26}$). B2B
was less sensitive than the backward regression (length: $t-value=33.6$,
$p=10^{-50}$, $t-value=31.7$, frequency:$p=10^{-48}$). However, as detailed in
the introduction, backward models confound collinear factors. Consequently,
the high sensitivity of our backward regression is likely to come from a loss of specificity. This possibility is corroborated by the fact that the dynamics
associated with word length and word frequency are remarkably similar in the
backward model.
